---
title: "TFM_AD"
output: html_document
date: "2025-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Import Libaries

```{r}
rm(list=ls())
gc()
source("./code/00.libraries.R")
source("./code/01.functions.R")

# Global Variables
output_path <- "output/"
image_number <- 1  # Initialize image counter for saving figures
# Create output directory if it doesn't exist
create_directories(output_path)

```


# 1. Introduction

# 2. Goals

# 3. M&M
For databases, we will start with two different datasetsÑ

- Syntetic databases with a plethora of variables commonly checked in AD
- Real MRI images downloaded from Kaggle (https://www.kaggle.com/datasets/abdullahtauseef2003/adni-4c-alzheimers-mri-classification-dataset/data) with 

- A4 Dataset
Anti-Amyloid Treatment in Asymptomatic Alzheimer's Disease
https://doi.org/10.1001/jamaneurol.2020.0387

https://discover.a4studydata.org/catalogue/studies/1901af97-b7cf-46ef-ab5c-b43212902b72

We will use A4 dataset with to perspectives:
- First, use tabular data to obtain a probability of AD in patients
- Second, use PET images to get the probability (or score) of Tau or B-amiloid deposits.
- Third, and most importatn, mix both CNN in a single one to obtain the mixed probability bases in tabular and images data.

First aim will be done in R, with EDA, remove outliers, impute missing values, and select variables. For ML based in tabular data, I will use R (tidymodels/H2O) and/or Python.
Second and third aim will be done in Pyton.

## 3.1. Loading tabular data
The database to be used comes from the A4 study. The data includes various clinical, demographic, and imaging variables.

### Screening

```{r}
## Sankey plot
# Create a data frame for the flows
flow_data <- data.frame(
  Visit1  = c("Screened", "Screened", "Screened"),
  Visit2  = c("Excluded", "Continued", "Continued"),
  Amyloid = c("Screen Failed",        "AB-",      "AB+"),
  Study   = c("Screen Failed",        "LEARN",    "A4"),
  n       = c(2277,      3163,       1323)
)

# Set desired top-to-bottom order
flow_data$Amyloid <- factor(as.character(flow_data$Amyloid), levels = c("AB+", "AB-", "Screen Failed"))
flow_data$Study   <- factor(as.character(flow_data$Study),   levels = c("A4", "LEARN", "Screen Failed"))

# Replace NA labels with something readable
flow_data$Amyloid <- fct_explicit_na(flow_data$Amyloid, na_level = "—")
flow_data$Study   <- fct_explicit_na(flow_data$Study,   na_level = "—")

# Plot
p <- ggplot(flow_data,
       aes(axis1 = Visit1, axis2 = Visit2, axis3 = Amyloid, axis4 = Study, y = n)) +
  scale_x_discrete(limits = c("Visit 1", "Visit 2", "Amyloid", "Study"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = Study), width = 1/12, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "grey90", color = "black") +
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\n(n=", after_stat(count), ")")),
            size = 3.5) +
  theme_minimal(base_size = 14) +
  theme(axis.title = element_blank(),
        panel.grid = element_blank()) +
  labs(title = "Participant Flow in A4/LEARN Screening")

print(p)
save_plot("participant_flow_sankey", p, height = 6)
rm(flow_data)
```

### Load CSV

```{r}
# Load all csv files in the directory Derived data

# List CSV file paths and names
file_paths <- list.files(path = "input_data/A4/Derived Data", pattern = "\\.csv$", full.names = TRUE)
file_names <- basename(file_paths) %>% str_remove("\\.csv$")

# Create an environment to store log
removal_log <- list()

# Function to read, convert chars to factors, remove 1-level factors, and log removals
read_and_clean <- function(file, name) {
  df <- read_csv(file, show_col_types = FALSE) %>%
    mutate(across(where(is.character), as.factor))
  
  # Identify factor columns with only one level
  one_level_cols <- names(df)[sapply(df, function(col) is.factor(col) && nlevels(col) == 1)]
  # Identify columns with zero SD (constant columns)
  zero_sd_cols <- names(df)[sapply(df, function(col) is.numeric(col) && sd(col, na.rm = TRUE) == 0)]
  # Combine both sets of columns to remove
  one_level_cols <- unique(c(one_level_cols, zero_sd_cols))
  # Log if any were found
  if (length(one_level_cols) > 0) {
    removal_log[[name]] <<- one_level_cols
  }
  
  # Remove those columns
  df %>% select(!all_of(one_level_cols))
}

# Read and clean all, storing in a named list
Derived_Data <- map2(file_paths, file_names, ~ read_and_clean(.x, .y)) %>%
  set_names(file_names)

# Show log with character columns removed, those with only one factor level
removal_log

save(Derived_Data, file = paste0(output_path, "RData/","Derived_Data.RData"))

# Summarize structure into a tidy format
structure_summary <- map_dfr(Derived_Data, ~ tibble(
  column = names(.x),
  class   = sapply(.x, class),
  n_levels = sapply(.x, function(col) if (is.factor(col)) nlevels(col) else NA),
  n_missing = sapply(.x, function(col) sum(is.na(col))),
  n_unique = sapply(.x, function(col) if (is.factor(col)) n_distinct(col) else length(unique(col))),
  n_rows = nrow(.x)
  ), .id = "source_file")
# Save the structure summary
write_csv(structure_summary, paste0(output_path, "tables/","structure_summary.csv"))
# Clean environment
rm(file_paths, file_names, removal_log)

```
#### Split ADQS df

```{r}
#Select test for every visit
test_QS <- Derived_Data$ADQS %>%
  select( SUBSTUDY, BID, VISITCD, VISIT, AVISIT, EPOCH, QSSEQ, QSTESTCD, QSTEST, QSVERSION, QSLANG, QSSTRESN, QSBLRES, QSCHANGE, QSBLSD)

#Save test data
write_csv(test_QS, paste0(output_path, "tables/","test_qs_data.csv"))

# Remove test variables form ADQS
Derived_Data$ADQS <- Derived_Data$ADQS %>%
  select(-c(QSSEQ, QSTESTCD, QSTEST, QSVERSION, QSLANG, QSSTRESN, QSBLRES, QSCHANGE, QSBLSD))

# Remove duplicates in ADQS
Derived_Data$ADQS <- Derived_Data$ADQS %>%
  distinct(BID, SUBSTUDY, VISITCD, VISIT, AVISIT, EPOCH, .keep_all = TRUE)
```


### Glosary of the data

```{r}
# Load glosary of data
glosary <- read_csv("input_data/A4/Documents/Data Dictionaries/derived_datadic.csv", show_col_types = FALSE)

# Clean glosary: remove empty rows and unnecessary columns
glosary <- glosary %>%
  select(FIELD_NAME, FIELD_FORMAT, FIELD_DESC, FIELD_CODE) %>% 
  #Remove duplicates
  distinct()


# Join FIELD_FORMAT, FIELD_DESC, FIELD_CODE with structure_summary
structure_summary <- structure_summary %>%
  left_join(glosary, by = c("column" = "FIELD_NAME")) %>%
  select(source_file, column, class, n_levels, n_missing, n_unique, n_rows, FIELD_FORMAT, FIELD_DESC, FIELD_CODE)

# Save the structure summary
write_csv(structure_summary, paste0(output_path, "tables/","structure_summary.csv"))
rm(glosary)
```

## Select longitudinal and cross-sectional data
Longitudinal data is defined as data that has multiple measurements for the same subject over time, while cross-sectional data is a snapshot of a single point in time.
Longitudinal data will change over time (VISIT, AVISIT, BID), while cross-sectional data will not change over time.
```{r}
# Función para identificar variables constantes y longitudinales por paciente
get_variable_type <- function(df) {
  # Nos aseguramos que la columna BID exista
  if (!"BID" %in% colnames(df)) return(NULL)
  
  # Eliminamos columnas vacías o todas NA
  df <- df %>% select(where(~ !all(is.na(.))))

  # Excluimos columnas no informativas
  exclude_cols <- c("BID", "SUBSTUDY")
  test_vars <- setdiff(colnames(df), exclude_cols)

  # Para cada variable, comprobamos si tiene solo un valor único por BID
  var_type <- map_dfr(test_vars, function(var) {
    tmp <- df %>%
      group_by(BID) %>%
      summarise(n_unique = n_distinct(.data[[var]], na.rm = TRUE), .groups = "drop") %>%
      summarise(max_unique = max(n_unique, na.rm = TRUE)) %>%
      mutate(variable = var,
             type = ifelse(max_unique == 1, "constant", "longitudinal"))
  })

  return(var_type)
}

# Aplicar la función a cada dataframe
variable_types <- map(Derived_Data, get_variable_type)


# Crear una lista con tipo constante o longitudinal y nombre del df
var_summary <- map2_dfr(
  variable_types,
  names(variable_types),
  ~ mutate(.x, dataset = .y)
)

var_summary <- var_summary %>%
  select(dataset, type, variable) %>%
  arrange(dataset,variable, )

# Save variable types
write_csv(var_summary, paste0(output_path, "tables/","variable_types.csv"))
save(variable_types, file = paste0(output_path, "RData/","variable_types.RData"))
```

### Cross-sectional data

```{r}
cross_sectional_df_list <- map2(Derived_Data, variable_types, function(df, types) {
  if (is.null(types)) return(NULL)
  constant_vars <- filter(types, type == "constant")$variable
  vars_to_select <- c("BID", "SUBSTUDY", constant_vars)
  df %>% 
    select(any_of(vars_to_select)) %>%
    distinct(BID, .keep_all = TRUE)
})

# 1. Filtrar sólo los data.frames válidos
cross_sectional_df_list_clean <- cross_sectional_df_list %>%
  discard(~ is.null(.x) || !is.data.frame(.x))

# 2. Asegurar que BID es carácter en todos los dataframes
cross_sectional_df_list_clean <- map(cross_sectional_df_list_clean, function(df) {
  df %>% mutate(BID = as.character(BID))
  
})

# 3. Hacer el merge seguro
cross_sectional_merged <- reduce(cross_sectional_df_list_clean, full_join, by = "BID")

# 2. Resolver columnas duplicadas: AGE.x, AGE.y, etc.
# Paso automático: identificar columnas con el mismo prefijo
consolidate_columns <- function(df) {
  base_names <- names(df)
  # Extraer columnas duplicadas por sufijos .x, .y, etc.
  name_stems <- gsub("\\..*$", "", base_names)
  duplicated_names <- name_stems[duplicated(name_stems)]
  unique_dups <- unique(duplicated_names)

  for (stem in unique_dups) {
    cols <- base_names[startsWith(base_names, paste0(stem, "."))]
    if (length(cols) > 0 && stem %in% names(df)) {
      cols <- c(stem, cols)  # incluir la original
    } else {
      cols <- base_names[name_stems == stem]
    }

    # Crear nueva columna coalescida
    df[[stem]] <- df %>%
      select(any_of(cols)) %>%
      purrr::pmap_chr(~ coalesce(!!!list(...)) %>% as.character())

    # Eliminar columnas duplicadas
    df <- df %>% select(-any_of(setdiff(cols, stem)))
  }
  return(df)
}

# 3. Aplicar consolidación
cross_sectional_df <- consolidate_columns(cross_sectional_merged)
# Remove previous BID from cross sectional data
previous_bid_c <- cross_sectional_df %>% 
  select(BID, PREVBID) %>%
  distinct()

# Remove previous BID from cross sectional data
cross_sectional_df <- cross_sectional_df %>%
  select(-PREVBID)
# Save cross-sectional data
write_csv(cross_sectional_df, paste0(output_path, "tables/","cross_sectional_data.csv"))
# Clean enviroment
rm(cross_sectional_df_list, cross_sectional_df_list_clean, cross_sectional_merged)
```

### Longitudinal data
```{r}

get_clean_longitudinal <- function(df_name, data_list, type_list) {
  if (df_name == "DS") return(NULL)  # omitir DS

  df <- data_list[[df_name]]
  types <- type_list[[df_name]]
  if (is.null(df) || is.null(types)) return(NULL)

  # Extraer variables longitudinales
  long_vars <- types %>%
    filter(type == "longitudinal") %>%
    pull(variable)

  # Variables de visita a retener si existen
  visit_vars <- c("VISIT", "VISITCD", "VISCODE", "AVISIT")

  df %>%
    select(any_of(c("BID", "SUBSTUDY", visit_vars, long_vars))) %>%
    mutate(across(everything(), ~ as.character(.)))
}

df_names <- setdiff(names(Derived_Data), "DS")  # omitir DS explícitamente

longitudinal_df_list <- map(df_names, get_clean_longitudinal, data_list = Derived_Data, type_list = variable_types) %>%
  compact()

longitudinal_df <- bind_rows(longitudinal_df_list)



# Obtener nombres de columnas base que tienen .z o .ch
all_cols <- colnames(longitudinal_df)

base_vars <- unique(gsub("\\.(z|ch)$", "", all_cols[grepl("\\.(z|ch)$", all_cols)]))
# Crear una copia para trabajar
longitudinal_df_clean <- longitudinal_df

for (base in base_vars) {
  col_z <- paste0(base, ".z")
  col_ch <- paste0(base, ".ch")
  col_base <- base
  
  # Crear columna consolidada si existe al menos una
  if (any(c(col_z, col_ch, col_base) %in% names(longitudinal_df_clean))) {
    new_col <- paste0(base, "_merged")
    
    longitudinal_df_clean[[new_col]] <- coalesce(
      longitudinal_df_clean[[col_z]],
      longitudinal_df_clean[[col_ch]],
      longitudinal_df_clean[[col_base]]
    )
  }
}

cols_to_remove <- unlist(lapply(base_vars, function(x) c(x, paste0(x, ".ch"), paste0(x, ".z"))))
cols_to_remove <- intersect(cols_to_remove, colnames(longitudinal_df_clean))

longitudinal_df <- longitudinal_df_clean %>% select(-all_of(cols_to_remove))


# Save longitudinal data
write_csv(longitudinal_df, paste0(output_path, "tables/","longitudinal_data.csv"))
# Clean environment
rm(df_names, longitudinal_df_list, longitudinal_df_clean, base_vars, all_cols, cols_to_remove)
```
### Remove previous patient ID




## Combine datasets in a single object

```{r}

# Combine cross-sectional and longitudinal data into a list
A4_data <- list(
  cross_sectional = cross_sectional_df,
  longitudinal = longitudinal_df,
  test_QS = test_QS,
  previous_bid = previous_bid_c
)
#source("./code/01.functions.R") 
A4_data$cross_sectional <- apply_column_classes(A4_data$cross_sectional, structure_summary)
A4_data$longitudinal_df <- apply_column_classes(A4_data$longitudinal, structure_summary)
A4_data$test_QS <- apply_column_classes(A4_data$test_QS, structure_summary)

#Save A4_data
save(A4_data, file = paste0(output_path, "RData/","A4_data.RData"))
rm(cross_sectional_df, longitudinal_df, test_QS, previous_bid_c)
```

## Load combined data

```{r}
load(paste0(output_path, "RData/","A4_data.RData"))
```

# Variables conversion to factors
## Cross-sectional data
```{r}
# Convert character columns to factors in cross-sectional data
A4_data$cross_sectional <- A4_data$cross_sectional %>%
  mutate(
    BID = as.character(BID),
    AGECAT = factor(case_when(
      AGEYR < 65 ~ "Age < 65",
      AGEYR >= 65 &
        AGEYR < 75 ~ "65 <= Age < 75",
      AGEYR >= 75 & AGEYR < 85 ~ "75 <= Age < 85",
      AGEYR >= 85 ~ "Age >= 85"
    )),
    SEX = factor(
      case_when(SEX == 1 ~ "Female", SEX == 2 ~ "Male"),
      levels = c("Male", "Female")
    ),
    RACE = factor(case_when(
      RACE == 1 ~ "White",
      RACE == 2 ~ "Black or African American",
      RACE == 58 ~ "Asian",
      RACE == 79 ~ "Native Hawaiian or Other Pacific Islander",
      RACE == 84 ~ "American Indian or Alaskan Native",
      RACE == 97 ~ "Unknown or Not Reported",
      RACE == 100 ~ "More than one race"
    )),
    MARITAL = factor(case_when(
      MARITAL == 2 ~ "Divorced",
      MARITAL == 4 ~ "Never married",
      MARITAL == 5 ~ "Widowed",
      MARITAL == 11 ~ "Married",
      MARITAL == 97 ~ "Unknown or Not Reported"
    )),
    ETHNIC = factor(case_when(
      ETHNIC == 50 ~ "Hispanic or Latino",
      ETHNIC == 56 ~ "Not Hispanic or Latino",
      ETHNIC == 97 ~ "Unknown or Not reported"
    )),
    ALCHLBL = factor(case_when(ALCHLBL == 0 ~ "No", ALCHLBL == 1 ~ "Yes")),
    CFBL = factor(case_when(CFBL == 0 ~ "No", CFBL == 1 ~ "Yes")),
    TBBL = factor(case_when(TBBL == 0 ~ "No", TBBL == 1 ~ "Yes")),
    WRKRET = factor(case_when(WRKRET == 1 ~ "Yes", WRKRET == 0 ~ "No", WRKRET == 96 ~ "Not Applicable")),
    APOEGNPRSNFLG = factor(case_when(APOEGNPRSNFLG == 1 ~ "Yes", APOEGNPRSNFLG == 0 ~ "No")),
    AGEYR = as.numeric(AGEYR),
    SUVRCER = as.numeric(SUVRCER),
    AMYLCENT = as.numeric(AMYLCENT),
    EDCCNTU = as.numeric(EDCCNTU),
    COGDSSTTSV6 = as.numeric(COGDSSTTSV6),
    COGLMDRTSV6 = as.numeric(COGLMDRTSV6),
    TX = factor(TX, levels = c("Placebo", "Solanezumab")),
    COMPLETER_label = factor(case_when(SUBJCOMPTR == 1 ~ "Completer", TRUE ~ "Dropout"))
  )


#Save A4_data
save(A4_data, file = paste0(output_path, "RData/","A4_data.RData"))

```


# 4. Results

## 4.1. EDA

```{r}
#Resumen de datos
skimr::skim(A4_data$cross_sectional)
```


```{r}
funModeling::status(A4_data$cross_sectional)
```


```{r}
funModeling::profiling_num(A4_data$cross_sectional)
```

```{r}
# Plot numeric variables distribution
plot_num <- function(data, vars = NULL, ncol = 4,
                     title = "Numeric Variables Distribution",
                     theme = theme_minimal()) {
  
  # Si no se especifican, usar todas las numéricas
  if (is.null(vars)) {
    vars <- data %>% select(where(is.numeric)) %>% colnames()
  }
  
  # Pivotar a formato largo
  df_long <- data %>%
    select(any_of(vars)) %>%
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
    filter(!is.na(value))

  # Gráfico
  p <- ggplot(df_long, aes(x = value)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black", alpha = 0.6) +
    geom_density(color = "red", linewidth = 0.1, adjust = 1.2) +
    facet_wrap(~variable, scales = "free", ncol = ncol) +
    labs(title = title, x = NULL, y = "Density") +
    theme

  print(p)
  save_plot("Numeric_variables_distribution", p, height = 20, width = 16)
}


numeric_var <- A4_data$cross_sectional %>%
  select(where(is.numeric)) %>%
  colnames()

plot_num(A4_data$cross_sectional, vars = numeric_var)
         
```


```{r}


plot_cat <- function(data, vars = NULL, ncol = 2,
                     title = "Categorical Variables Distribution",
                     theme = theme_minimal()) {
  
  # Si no se especifican, usar todas las categóricas
  if (is.null(vars)) {
    vars <- data %>%
      select(where(~ is.character(.) || is.factor(.))) %>%
      colnames()
  }
  
  # Formato largo
  df_long <- data %>%
    select(any_of(vars)) %>%
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
    filter(!is.na(value)) %>%
    mutate(value = as.factor(value))

  # Plot
  p <- ggplot(df_long, aes(x = value)) +
    geom_bar(fill = "steelblue", color = "black", alpha = 0.7) +
    facet_wrap(~variable, scales = "free", ncol = ncol) +
    labs(title = title, x = NULL, y = "Count") +
    theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p)
  
  save_plot("Categorical_variables_distribution", p, height = 20, width = 16)
}

categorical_vars <- A4_data$cross_sectional %>%
  select(where(~ is.factor(.))) %>%
  colnames()

plot_cat(A4_data$cross_sectional, vars = categorical_vars)

```
```{r}

for (i in categorical_vars) {
  plotar(A4_data$cross_sectional, input= numeric_var, target= i, plot_type = "boxplot", path_out = paste0(output_path, "figures/"))
}

```

##### Remove LEARN and SF Data #####################################


## 4.1. Database generation

### 4.1.1. Preprocessing





### 4.1.2. EDA

### 4.1.2. Tidymodels

## 4.2. Images databases

### 4.2.1. Image classification

### 4.2.2. Data Augmentation

### 4.2.3. ML in Python


## 4.2. ML integration

# 5. Discussion

# 6. Bibliography

# 7. sessionInfo
```{r}
sessionInfo()
```

